---
title: "capstone_statistical_analysis"
output: html_document
---

***

```{r, echo = FALSE, message = FALSE}
library(xts)
data <- read.csv("levels2.csv", header = TRUE)
data$dd.mmm.yy <- as.Date(data$dd.mmm.yy, "%d-%b-%y")
ts_df <- xts(data[, 3:7], order.by = data[, 2])
ts_df_diff <- ts_df / lag(ts_df, 1) - 1
ts_df_diff <- ts_df_diff[2:nrow(ts_df_diff), ]
```

After we import the data and coerce it to a time series dataset via xts, we first quickly check the range of each level:

```{r, echo = FALSE}
range_matrix <- as.matrix(apply(ts_df, 2, range))
rownames(range_matrix) <- c("Low", "High")
range_matrix
```

We also check the quantiles for each:

```{r, echo = FALSE}
apply(ts_df, 2, quantile)
```

After we first-difference the time series, we again check the range and quantiles:

```{r, echo = FALSE}
range_matrix_diff <- as.matrix(apply(ts_df_diff, 2, range))
rownames(range_matrix_diff) <- c("Low", "High")
range_matrix_diff
apply(ts_df_diff, 2, quantile)
```

We now plot the five time series:

```{r, echo = FALSE, message = FALSE}
library(tidyr)
library(ggplot2)
df <- as.data.frame(ts_df)
df$Date <- rownames(df)
df$Date <- as.Date(df$Date, foramt = "%Y-%m-%d")
df_tall <- df %>% gather(key = "Country_Region", value = "Level", -Date)
ggplot(df_tall, aes(x = Date, y = Level, col = Country_Region)) + geom_point() + facet_wrap(. ~ Country_Region, ncol = 5) + scale_x_date(date_labels = "%Y", date_breaks = "6 years")
```

We now fit a linear trend to each of the time series:

```{r, echo = FALSE}
ggplot(df_tall, aes(x = Date, y = Level, col = Country_Region)) + geom_point() + geom_smooth(method = "lm", se = TRUE, col = "black") + facet_wrap(. ~ Country_Region, ncol = 5) + scale_x_date(date_labels = "%Y", date_breaks = "6 years")
```

It appears as if a linear model will not effectively capture the dynamics of any of the time series.

We now plot a histogram for each of the quarterly returns series:

```{r, echo = FALSE, message = FALSE}
df_diff <- as.data.frame(ts_df_diff)
df_diff$Date <- rownames(df_diff)
df_diff$Date <- as.Date(df_diff$Date, foramt = "%Y-%m-%d")
df_diff_tall <- df_diff %>% gather(key = "Country_Region", value = "Level", -Date)
ggplot(df_diff_tall, aes(x = Level, fill = Country_Region)) + geom_histogram() + facet_wrap(. ~ Country_Region, ncol = 5)
```

We now check to see how well the quarterly returns can be modeled by a Gaussian distribution:

```{r, echo = FALSE}
KS <- apply(ts_df_diff, 2, ks.test, "pnorm")
KS_pvalues <- unlist(lapply(KS, '[[', 2))
KS_pvalues <- as.matrix(t(KS_pvalues))
rownames(KS_pvalues) <- "P value"
KS_pvalues
```

Applying the Kolmogorov-Smirnov test suggests that the returns for all of the time series do not follow a Gaussian distribution, and hence we must take a measured approach we any further exploration.
