---
title: "Capstone Data Wrangling"
output: html_document
---

***

```{r, echo = FALSE}
data <- read.csv("levels2.csv", header = TRUE)
```

The dataset ranges from `r data[1,2]` to `r data[nrow(data), 2]`.  Given that we are only concerned with end-of-quarter data in order to align the periodicity, we remove the last observation, in which case the dataset ranges to `r data[nrow(data) - 1, 2]`.

```{r, echo = FALSE}
data <- data[1:nrow(data)-1, ]
```

When we check the structure of the dataset:

```{r, echo = FALSE}
str(data)
```

We see that the full-date column is of type "factor," so we are best to coerce it to a date `r data$dd.mmm.yy <- as.Date(data$dd.mmm.yy, format = "%d-%b-%y")`and then re-check the structure:

```{r, echo = FALSE}
str(data$dd.mmm.yy)
```

A check of the dataset reveals that there are `r sum(is.na(data))` NA's.

We verify that the days in the dataset are all mapped to end-of-month: `r sort(unique(as.numeric(format(data$dd.mmm.yy, "%d"))))`

And that the months are quarterly: `r unique(as.numeric(format(data$dd.mmm.yy, "%m")))`

And we see that every quarterly date is accounted for in the time series:

```{r, echo = FALSE}
table(as.numeric(format(data$dd.mmm.yy, "%Y")), as.numeric(format(data$dd.mmm.yy, "%m")))
```

We are now ready to proceed with statistical analysis of our dataset.